Determinism, Complexity, and Predictability in Computer Performance

### 2014-07-16 model.tex
second round of LizB edits

\item The \emph{ARIMA} (auto-regressive integrated moving average)
  method, a common linear forecast strategy, and

\item The \emph{LMA} (Lorenz method of analogues) method, which uses a
  near-neighbor forecast strategy on a dynamical reconstruction of the
  signal.
\end{itemize}
ARIMA is based on standard autoregressive techniques.  LMA is designed
to capture and exploit the deterministic structure of a signal from a
nonlinear dynamical system.  The \naive ~and random-walk methods,
somewhat surprisingly, often outperform these more-sophisticated
prediction strategies in the case of highly complex signals, as
discussed below.

@@

Of course, methods that could capture and exploit the geometry of the
data, or its temporal patterns, would be far more effective in the
situations described in the previous paragraph.  The ARIMA and LMA
methods introduced in Sections~\ref{sec:arima} and~\ref{sec:lma} are
designed to do exactly that.
% Note, though, that periodic patterns
%appear only in signals that are at the low end of the %complexity
%spectrum.
However, if a signal contains little predictive structure, forecast
strategies like ARIMA and LMA have nothing to work with and thus will
often be outperformed by the two simple strategies described in this
section.  This effect is explored further in Sections~\ref{sec:accuracy}
and~\ref{sec:results}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% here is the old version of this material, for use in other documents...
%
%Perhaps the simplest way to capture and exploit the structure of data
%is to fit a hyperplane to the dataset and then use that it to forecast
%new data points.  The roots of this date back to Yule's 1927 invention
%of the autoregressive schema~\cite{weigend93}, which forecasts the
%next time step through a weighted average of past observations: $$p_i
%= \phi(B^p)x_{i}$$ where $\phi(\cdot)$ is a polynomial of degree $p$
%that is fit to the past $p+1$ [[???]] values of $x_i$ via the
%backshift operator $B$ \footnote{I changed the superscript to $k$ to
%  make this definition general---not connected to any of the terms
%  that have specific meanings later in this section.}:
%$$B^k x_i = x_{i-k}$$
%%
%with $k=1...p$.  To account for noise in the data, one can add a
%so-called ``moving average'' term to the model:
%$\theta(B^q)\epsilon_i$, where $\theta(\cdot)$ is a polynomial of
%degree $q$ and $\epsilon$ is white noise with the same mean and
%variance as the data.
%% if you leave this in, you'll need to explain a lot about the data
%% and the calculation
%% $\{\epsilon_i\}\sim WN(0,\sigma^2)$.
%To remove nonstationarities in the data, one can detrend it using a
%differencing operation: $(1-B^d)x_i$.
%
%A strategy that incorporates all three of these features is called a
%\emph{nonseasonal ARIMA} model of order $(p,d,q)$:
% $$\phi(B^p)(1-B^d) x_i = \theta(B^q)\epsilon_i$$
%%
%{\color{red} You had superscripts on some of the $B$s and not on
%  others.  I added them but I may have gotten them wrong.  I also
%  changed some $(1-B)^d$s to $(1-B^d)$s to make things consistent with
%  the first use of that construction.  Please make sure things are
%  consistent, here and throughout this section.}  Here, $p$, $d$ and
%$q$ correspond to the orders of the autoregressive, detrending, and
%moving average terms, respectively.
%
%If periodic structure is present in the data, a \emph{seasonal ARIMA}
%model of order $(p,d,q)(P,D,Q)$ can be a good choice:
%%
%$$\Phi(B^P)\phi(B^p)(1-B^m)^D(1-B^d) x_i =
%\Theta(B^Q)\theta(B^q)\epsilon_i$$
%%
%Here $\Phi(\cdot)$ and $\Theta(\cdot)$ are polynomials of degree $P$
%and $Q$, respectively [[which accomplish what purpose?  why?  how?
%    where did $D$ come from and what does it do?]].  $m$ is the
%seasonal frequency [[shouldn't this be ``period''?]].  [[Explain why
%    one should choose $P=Q=m$, since that's what the equation below
%    implicitly does.]]  To use a seasonal ARIMA model, one first uses
%the detrending term to remove nonstationarity: $$\hat{x_i} =
%(1-B^m)^D(1-B^d) x_{i}$$ and then creates the forecast $p_i$ by
%evaluating $$\Phi(B^m)\phi(B^p)\hat{X_i} =
%\Theta(B^m)\theta(B^q)\epsilon_i$$
% ...down to here
%%%%%%%%%%%%%%%%%%%%%%%%

@@

There is a vast amount of theory and literature regarding the
construction and use of models of this type; we refer the reader to
\cite{davislinearts} for an in-depth exploration.  For the purposes of
this paper, where the goal is to explore the relationship between
predictability and complexity across a broad array of forecast
strategies, seasonal ARIMA models are a good exemplar of the class of
linear predictors.  Fitting such a model to a dataset involves
choosing values for the various free parameters in the autoregressive,
detrending, moving average, and filtering terms.  We employ the
automated fitting techniques described in~\cite{autoARIMA} to
accomplish this.  This procedure uses sophisticated methods---KPSS
unit-root tests~\cite{KPSSunit}, a customization of the Canova-Hansen
test~\cite{Canova1995}, and the Akaike information
criterion~\cite{akaike1974}, conditioned on the maximum likelihood
of the model fitted to the detrended data---to select good values for
the free parameters of the ARIMA model.

ARIMA forecasting is a common and time-tested procedure.  Its
adjustments for seasonality, nonstationarity, and noise make it an
appropriate choice for short-term predictions of time-series data
generated by a wide range of processes.  If information is being
generated and/or transmitted in a nonlinear way, however, a global
linear fit is inappropriate and ARIMA forecasts can be inaccurate.
Another weakness of this method is prediction horizon: an ARIMA
forecast is guaranteed to converge to the mean after some number of
predictions, depending on model order.  To sidestep this issue, we
build forecasts in a stepwise fashion: i.e., fit the model to the
existing data, use that model to perform a one-step prediction, rebuild it
using the latest observations, and iterate until the desired
prediction horizon is reached.  (For consistency, we take the same
approach with the other three models in this study as well, even
though doing so amounts to artificially hobbling LMA.)

@@

  \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colRWForecast}
    \caption{\col\\ random walk }
    \label{fig:colRW}
  \end{subfigure}%
   \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/gccRWForecast}
    \caption{\gcc\\ random walk }
    \label{fig:gccRW}
  \end{subfigure}%
     \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/svdfiveRWForecast}
    \caption{\svdfive\\ random walk}
    \label{fig:svd5RW}
  \end{subfigure}%
  \\
      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colMeanForecast}
    \caption{\col\\ na\"ive }
    \label{fig:colMEAN}
  \end{subfigure}%
   \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/gccMeanForecast}
    \caption{\gcc\\ na\"ive }
    \label{fig:gccMEAN}
  \end{subfigure}%
     \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/svdfiveMeanForecast}
    \caption{\svdfive\\ na\"ive }
    \label{fig:svd5MEAN}
  \end{subfigure}%
  \\
    \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colARIMAForecast}
    \caption{\col \\ ARIMA}
    \label{fig:colARIMA}
  \end{subfigure}
  \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/gccARIMAForecast}
    \caption{\gcc \\ ARIMA }
    \label{fig:gccARIMA}
  \end{subfigure}%
  \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/svdfiveARIMAForecast}
    \caption{\svdfive \\ ARIMA }
    \label{fig:svd5ARIMA}
  \end{subfigure}%
  \\

@@

\begin{table*}
  \renewcommand{\arraystretch}{1.25}
  \begin{tabular}{lccc}
    & {\large \col} & {\large \gcc} & {\large \svdfive} \\
    & & & \\
    \begin{sideways}\large random walk\end{sideways} &
    \includegraphics[width=0.6\columnwidth]{figs/colRWForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/gccRWForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/svdfiveRWForecast} \\
    \begin{sideways}\large na\"ive\end{sideways} &
    \includegraphics[width=0.6\columnwidth]{figs/colMeanForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/gccMeanForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/svdfiveMeanForecast} \\
    \begin{sideways}\large ARIMA\end{sideways} &
    \includegraphics[width=0.6\columnwidth]{figs/colARIMAForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/gccARIMAForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/svdfiveARIMAForecast} \\
    \begin{sideways}\large LMA\end{sideways} &
    \includegraphics[width=0.6\columnwidth]{figs/colLMAForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/gccLMAForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/svdfiveLMAForecast}
  \label{fig:forecast-example}
  \end{tabular}
  \caption{Predicted ($p_j$) versus true values ($c_j$) for forecasts of \col,
     \gcc, and \svdfive and all four prediction strategies.
  }
  \label{fig:forecast-example}
\end{table*}
In these images, the vertical axis is the prediction $p_j$ and the
horizontal axis is the true continuation $c_j$.  On such a plot, a
perfect prediction would lie on the diagonal.  LMA, for instance,
generates a very accurate prediction of the \col data, while ARIMA
does not.  Horizontal lines result when a constant predictor (e.g.,
\naive) is used on a non-constant signal.  Point clouds reflect the
structure of the distribution of the errors---roughly normally
distributed, for example, in the case of LMA on \gcc.  Note that the
shapes of Figures~\ref{fig:colRW} and~\ref{fig:colARIMA} are
reminiscent of the projected embedding in Figure~\ref{fig:embedding}a.
For a random-walk predictor, a $p_j$ vs. $c_j$ plot is equivalent to a
two-dimensional embedding with $\tau=1$.  For ARIMA, the
correspondence is not quite as simple, since the $p_j$ values are
linear combinations of a number of past values of the $c_j$, but the
effect is largely the same\footnote{The temporal \emph{ordering} of
  the points on the ARIMA $p_j$ vs. $c_j$ plot does not match that of
  a projected embedding of the time series, however.}.

@@

MASE scores for all 360 experiments are tabulated in the three
middle columns in Table~\ref{tab:error}.
\begin{table*}
\caption{Mean
absolute scaled error (MASE) scores and weighted permutation entropies for all eight
   processes studied in this paper}
  \begin{center}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill} } ccccc}
  \hline\hline Signal & LMA MASE  & ARIMA MASE & na\"{i}ve MASE  & Weighted Permutation Entropy \\
  \hline
 \col           & $ 0.050 \pm0.002  $ & $0.599  \pm 0.211 $ & $0.571\pm0.002$&  $0.513 \pm 0.003$ \\

@@

\svdfive     & $ 0.718\pm 0.048 $ & $2.370  \pm 0.051 $ & $20.870 \pm 0.192$&  $0.678 \pm 0.007$ \\
 \svdsix     & $ 0.739\pm 0.068 $ & $ 1.438\pm 0.061$ & $2.197\pm0.083$&  $0.748 \pm 0.011$ \\
  \hline\hline
  \end{tabular*}
  \end{center}
 \label{default}
 \label{tab:error}
  \end{table*}%
Comparing these with the geometry of the plots in
Figure~\ref{fig:forecast-example}, one can see some obvious
correspondences.  The average LMA MASE score for the \col signals
was $0.050$, for instance, while ARIMA scored much worse ($0.599$).
That is, LMA performed roughly 20 times better on \col signals than a
random-walk predictor, while ARIMA only outperformed random walk by a
factor of 1.7.  This is in accordance with the visual appearance of
the corresponding images in Figure~\ref{fig:forecast-example}.  While
the correspondence between MASE score and the visual appearance of
these kinds of plots is clear cut in this case, that is not always so.
The plots of LMA predictions of \col and \svdfive both lie near the
diagonal, for instance, but the corresponding MASE scores were very
different: $0.050$ for \col and $0.718$ for \svdfive (resp., 20 and
1.4 times better than random-walk forecasts of the same signals).
This is a result of the normalization in the MASE score calculation.
Recall from Section~\ref{sec:simple} that the random-walk method
performs especially poorly on signals that oscillate rapidly.  \col
fits this description, so the random-walk error on this signal---the
denominator of the three MASE scores in the first row of the
Table---is pathologically high, which skews those scores down.
Random-walk prediction is very effective for the \svdfive signal, on
the other hand, so the denominator is small and the MASE scores are
skewed upwards.  Normalization is, as is often the case, a double-edged
sword.  Here, it facilitates comparisons across signals of different
types and lengths, but its particular pathologies must be taken into
account, as discussed at more length in Section~\ref{sec:results}.

% Table \ref{tab:error} provides the MASE distributions
% {\color{red}[[Joshua: Ryan, Is this the right word? we give mean $\pm$
%       std. dev but some have very skewed right tails]]} for each of
% the eight signals and three prediction strategies, these are averaged
% over 15 runs of each signal-method combination. For later comparison,
% Table \ref{tab:error} also has the complexity measure we introduce in
% Section \ref{sec:meaComplex}.

%[[and cherry pick a few examples of \gcc and \col to put in the text ]]

%---works quite well on the trace in Figure~%\ref{fig:ipc}, as%
%shown in Figure~\ref{fig:lmacol}.
%
%\begin{figure}
%   \centering
%\begin{subfigure}{\columnwidth}
%    \includegraphics[width=\columnwidth]{%colPredShortTS}
%    \caption{\col }
%    \label{fig:lmacol}
%  \end{subfigure}%
%
%    \begin{subfigure}{\columnwidth}
%    \includegraphics[width=\columnwidth]{%colPredShortTS}
%    \caption{\gcc}
%    \label{fig:lmagcc}
%  \end{subfigure}
%  \begin{subfigure}{\columnwidth}
%%    \includegraphics[width=\columnwidth]{%colPredShortTS}
%    \caption{\svdfive}
%    \label{fig:lmasvdfive}
%  \end{subfigure}%
%  \caption{{\color{red} [[If Liz thinks we should %include these I actually need to generate this %figure]]}An LMA-based forecast of
%       processor-efficiency performance trace from the%\col ,\gcc and \svdfive generating processes.  Red %circles and blue $\times$s are the true and
%       predicted values, respectively; vertical bars %show where these
%       values differ.}\label{fig:lmapredictions}
% \end{figure}
%
%\begin{figure}[htbp]
%  \centering
%   \includegraphics[width=\textwidth]{colPredShortTS}
%    \caption{A forecast of the last 4,000 points of the %signal in
%      Figure~\ref{fig:ipc} using an LMA-based strategy on %the
%      embedding in Figure~\ref{fig:embedding}.  Red circles %and blue
%      $\times$s are the true and predicted values, %respectively;
%      vertical bars show where these values differ.}
%\label{fig:lmacol}
%\end{figure}
%
%\begin{figure}[htbp]
%  \centering
%    \includegraphics[width=\textwidth]{colPredShortTS}
%     \caption{{\color{red} [[actually need to generate this %figure]]}An LMA-based forecast of the last 4,000 points of a
%       processor-load performance trace from the \gcc
%       benchmark.  Red circles and blue $\times$s are the %true and
%       predicted values, respectively; vertical bars show %where these
%       values differ.}
%\label{fig:lmagcc}
%\end{figure}
%
%On the other hand, if we use the same approach to %forecast the
%processor efficiency (IPC) of the \gcc time series, %the
%prediction is far less accurate; see Figure~%\ref{fig:lmagcc}.
%Figure~\ref{fig:lmasvdfive} is clearly a better %prediction than Figure~\ref{fig:gccLMA} but not nearly as good as the \col forecast.

>>>>

\item The \emph{ARIMA} (auto-regressive integrated moving average)
  method, a common linear forecast strategy, built using the
  \emph{AutoARIMA} procedure \cite{arima}, and

\item The \emph{LMA} (Lorenz method of analogues) method, which uses a
  near-neighbor forecast strategy on a dynamical reconstruction of the
  signal.
\end{itemize}
ARIMA models are based on standard autoregressive techniques.  LMA is
designed to capture and exploit the deterministic structure of a
signal from a nonlinear dynamical system.  The \naive ~and random-walk
methods, somewhat surprisingly, often outperform these
more-sophisticated prediction strategies in the case of highly complex
signals, as discussed below.

@@

Of course, methods that could capture and exploit the geometry of the
data, or its temporal patterns, would be far more effective in the
situations described in the previous paragraph.  The AutoARIMA and LMA
methods introduced in Sections~\ref{sec:arima} and~\ref{sec:lma} are
designed to do exactly that.
% Note, though, that periodic patterns
%appear only in signals that are at the low end of the %complexity
%spectrum.
However, if a signal contains little predictive structure, forecast
strategies like ARIMA and LMA have nothing to work with and thus will
often be outperformed by the two simple strategies described in this
section.  This effect is explored further in Sections~\ref{sec:accuracy}
and~\ref{sec:results}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% here is the old version of this material, for use in other documents...
%
%Perhaps the simplest way to capture and exploit the structure of data
%is to fit a hyperplane to the dataset and then use that it to forecast
%new data points.  The roots of this date back to Yule's 1927 invention
%of the autoregressive schema~\cite{weigend93}, which forecasts the
%next time step through a weighted average of past observations: $$p_i
%= \phi(B^p)x_{i}$$ where $\phi(\cdot)$ is a polynomial of degree $p$
%that is fit to the past $p+1$ [???] values of $x_i$ via the
%backshift operator $B$ \footnote{I changed the superscript to $k$ to
%  make this definition general---not connected to any of the terms
%  that have specific meanings later in this section.}:
%$$B^k x_i = x_{i-k}$$
%%
%with $k=1...p$.  To account for noise in the data, one can add a
%so-called ``moving average'' term to the model:
%$\theta(B^q)\epsilon_i$, where $\theta(\cdot)$ is a polynomial of
%degree $q$ and $\epsilon$ is white noise with the same mean and
%variance as the data.
%% if you leave this in, you'll need to explain a lot about the data
%% and the calculation
%% $\{\epsilon_i\}\sim WN(0,\sigma^2)$.
%To remove nonstationarities in the data, one can detrend it using a
%differencing operation: $(1-B^d)x_i$.
%
%A strategy that incorporates all three of these features is called a
%\emph{nonseasonal ARIMA} model of order $(p,d,q)$:
% $$\phi(B^p)(1-B^d) x_i = \theta(B^q)\epsilon_i$$
%%
%{\color{red} You had superscripts on some of the $B$s and not on
%  others.  I added them but I may have gotten them wrong.  I also
%  changed some $(1-B)^d$s to $(1-B^d)$s to make things consistent with
%  the first use of that construction.  Please make sure things are
%  consistent, here and throughout this section.}  Here, $p$, $d$ and
%$q$ correspond to the orders of the autoregressive, detrending, and
%moving average terms, respectively.
%
%If periodic structure is present in the data, a \emph{seasonal ARIMA}
%model of order $(p,d,q)(P,D,Q)$ can be a good choice:
%%
%$$\Phi(B^P)\phi(B^p)(1-B^m)^D(1-B^d) x_i =
%\Theta(B^Q)\theta(B^q)\epsilon_i$$
%%
%Here $\Phi(\cdot)$ and $\Theta(\cdot)$ are polynomials of degree $P$
%and $Q$, respectively [which accomplish what purpose?  why?  how?
%    where did $D$ come from and what does it do?].  $m$ is the
%seasonal frequency [shouldn't this be ``period''?].  [Explain why
%    one should choose $P=Q=m$, since that's what the equation below
%    implicitly does.]  To use a seasonal ARIMA model, one first uses
%the detrending term to remove nonstationarity: $$\hat{x_i} =
%(1-B^m)^D(1-B^d) x_{i}$$ and then creates the forecast $p_i$ by
%evaluating $$\Phi(B^m)\phi(B^p)\hat{X_i} =
%\Theta(B^m)\theta(B^q)\epsilon_i$$
% ...down to here
%%%%%%%%%%%%%%%%%%%%%%%%

@@

There is a vast amount of theory and literature regarding the
construction and use of models of this type; we refer the reader to
\cite{davislinearts} for an in-depth exploration.  For the purposes of
this paper, where the goal is to explore the relationship between
predictability and complexity across a broad array of forecast
strategies, seasonal ARIMA models are a good exemplar of the class of
linear predictors.  Fitting such a model to a dataset involves
choosing values for the various free parameters in the autoregressive,
detrending, moving average, and filtering terms.  We employ the
automated fitting techniques described in~\cite{autoARIMA} to
accomplish this, producing what we call an \emph{AutoARIMA model}.
This procedure uses sophisticated methods---KPSS unit-root
tests~\cite{KPSSunit}, a customization of the Canova-Hansen
test~\cite{Canova1995}, and the Akaike information
criterion~\cite{akaike1974}, conditioned on the maximum likelihood of
the model fitted to the detrended data---to select good values for the
free parameters of the AutoARIMA model.

ARIMA forecasting is a common and time-tested procedure.  Its
adjustments for seasonality, nonstationarity, and noise make it an
appropriate choice for short-term predictions of time-series data
generated by a wide range of processes.  If information is being
generated and/or transmitted in a nonlinear way, however, a global
linear fit is inappropriate and ARIMA forecasts can be inaccurate.
Another weakness of this method is prediction horizon: an ARIMA
forecast is guaranteed to converge to the mean after some number of
predictions, depending on model order.  To sidestep this issue, we
build forecasts in a stepwise fashion: i.e., fit the AutoARIMA model
to the existing data, use that model to perform a one-step prediction,
rebuild it using the latest observations, and iterate until the
desired prediction horizon is reached.  (For consistency, we take the
same approach with the other three models in this study as well, even
though doing so amounts to artificially hobbling LMA.)

@@

  \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colRWForecast}
    \caption{\col\\ random walk }
    \label{fig:colRW}
  \end{subfigure}%
   \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/gccRWForecast}
    \caption{\gcc\\ random walk }
    \label{fig:gccRW}
  \end{subfigure}%
     \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/svdfiveRWForecast}
    \caption{\svdfive\\ random walk}
    \label{fig:svd5RW}
  \end{subfigure}%
  \\
      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colMeanForecast}
    \caption{\col\\ na\"ive }
    \label{fig:colMEAN}
  \end{subfigure}%
   \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/gccMeanForecast}
    \caption{\gcc\\ na\"ive }
    \label{fig:gccMEAN}
  \end{subfigure}%
     \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/svdfiveMeanForecast}
    \caption{\svdfive\\ na\"ive }
    \label{fig:svd5MEAN}
  \end{subfigure}%
  \\
    \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colARIMAForecast}
    \caption{\col \\ AutoARIMA}
    \label{fig:colARIMA}
  \end{subfigure}
  \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/gccARIMAForecast}
    \caption{\gcc \\ AutoARIMA }
    \label{fig:gccARIMA}
  \end{subfigure}%
  \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/svdfiveARIMAForecast}
    \caption{\svdfive \\ AutoARIMA }
    \label{fig:svd5ARIMA}
  \end{subfigure}%
  \\

@@

\begin{table*}
  \renewcommand{\arraystretch}{1.25}
  \begin{tabular}{lccc}
    & {\large \col} & {\large \gcc} & {\large \svdfive} \\
    & & & \\
    \begin{sideways}\large random walk\end{sideways} &
    \includegraphics[width=0.6\columnwidth]{figs/colRWForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/gccRWForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/svdfiveRWForecast} \\
    \begin{sideways}\large na\"ive\end{sideways} &
    \includegraphics[width=0.6\columnwidth]{figs/colMeanForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/gccMeanForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/svdfiveMeanForecast} \\
    \begin{sideways}\large AutoARIMA\end{sideways} &
    \includegraphics[width=0.6\columnwidth]{figs/colARIMAForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/gccARIMAForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/svdfiveARIMAForecast} \\
    \begin{sideways}\large LMA\end{sideways} &
    \includegraphics[width=0.6\columnwidth]{figs/colLMAForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/gccLMAForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/svdfiveLMAForecast}
  \label{fig:forecast-example}
  \end{tabular}
  \caption{Predicted ($p_j$) versus true values ($c_j$) for forecasts of \col,
     \gcc, and \svdfive and all four prediction strategies.
  }
  \label{fig:forecast-example}
\end{table*}
In these images, the vertical axis is the prediction $p_j$ and the
horizontal axis is the true continuation $c_j$.  On such a plot, a
perfect prediction would lie on the diagonal.  LMA, for instance,
generates a very accurate prediction of the \col data, while AutoARIMA
does not.  Horizontal lines result when a constant predictor (e.g.,
\naive) is used on a non-constant signal.  Point clouds reflect the
structure of the distribution of the errors---roughly normally
distributed, for example, in the case of LMA on \gcc.  Note that the
shapes of Figures~\ref{fig:colRW} and~\ref{fig:colARIMA} are
reminiscent of the projected embedding in Figure~\ref{fig:embedding}a.
For a random-walk predictor, a $p_j$ vs. $c_j$ plot is equivalent to a
two-dimensional embedding with $\tau=1$.  For AutoARIMA, the
correspondence is not quite as simple, since the $p_j$ values are
linear combinations of a number of past values of the $c_j$, but the
effect is largely the same\footnote{The temporal \emph{ordering} of
  the points on the AutoARIMA $p_j$ vs. $c_j$ plot does not match that of
  a projected embedding of the time series, however.}.

@@

MASE scores for all 360 experiments are tabulated in the three
middle columns in Table~\ref{tab:error}.
\begin{table*}
\caption{Mean
absolute scaled error (MASE) scores and weighted permutation entropies for all eight
   processes studied in this paper}
  \begin{center}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill} } ccccc}
  \hline\hline Signal & LMA MASE  & AutoARIMA MASE & na\"{i}ve MASE  & Weighted Permutation Entropy \\
  \hline
 \col           & $ 0.050 \pm0.002  $ & $0.599  \pm 0.211 $ & $0.571\pm0.002$&  $0.513 \pm 0.003$ \\

@@

\svdfive     & $ 0.718\pm 0.048 $ & $2.370  \pm 0.051 $ & $20.870 \pm 0.192$&  $0.678 \pm 0.007$ \\
 \svdsix     & $ 0.739\pm 0.068 $ & $ 1.438\pm 0.061$ & $2.197\pm0.083$&  $0.748 \pm 0.011$ \\
  \hline\hline
  \end{tabular*}
  \end{center}
 \label{default}
 \label{tab:error}
  \end{table*}%
Comparing these with the geometry of the plots in
Figure~\ref{fig:forecast-example}, one can see some obvious
correspondences.  The average LMA MASE score for the \col signals was
$0.050$, for instance, while AutoARIMA scored much worse ($0.599$).
That is, LMA performed roughly 20 times better on \col signals than a
random-walk predictor, while AutoARIMA only outperformed random walk
by a factor of 1.7.  This is in accordance with the visual appearance
of the corresponding images in Figure~\ref{fig:forecast-example}.
While the correspondence between MASE score and the visual appearance
of these kinds of plots is clear cut in this case, that is not always
so.  The plots of LMA predictions of \col and \svdfive both lie near
the diagonal, for instance, but the corresponding MASE scores were
very different: $0.050$ for \col and $0.718$ for \svdfive (resp., 20
and 1.4 times better than random-walk forecasts of the same signals).
This is a result of the normalization in the MASE score calculation.
Recall from Section~\ref{sec:simple} that the random-walk method
performs especially poorly on signals that oscillate rapidly.  \col
fits this description, so the random-walk error on this signal---the
denominator of the three MASE scores in the first row of the
Table---is pathologically high, which skews those scores down.
Random-walk prediction is very effective for the \svdfive signal, on
the other hand, so the denominator is small and the MASE scores are
skewed upwards.  Normalization is, as is often the case, a
double-edged sword.  Here, it facilitates comparisons across signals
of different types and lengths, but its particular pathologies must be
taken into account, as discussed at more length in
Section~\ref{sec:results}.

% Table \ref{tab:error} provides the MASE distributions
% {\color{red}[Joshua: Ryan, Is this the right word? we give mean $\pm$
%       std. dev but some have very skewed right tails]} for each of
% the eight signals and three prediction strategies, these are averaged
% over 15 runs of each signal-method combination. For later comparison,
% Table \ref{tab:error} also has the complexity measure we introduce in
% Section \ref{sec:meaComplex}.

%[and cherry pick a few examples of \gcc and \col to put in the text ]

%---works quite well on the trace in Figure~%\ref{fig:ipc}, as%
%shown in Figure~\ref{fig:lmacol}.
%
%\begin{figure}
%   \centering
%\begin{subfigure}{\columnwidth}
%    \includegraphics[width=\columnwidth]{%colPredShortTS}
%    \caption{\col }
%    \label{fig:lmacol}
%  \end{subfigure}%
%
%    \begin{subfigure}{\columnwidth}
%    \includegraphics[width=\columnwidth]{%colPredShortTS}
%    \caption{\gcc}
%    \label{fig:lmagcc}
%  \end{subfigure}
%  \begin{subfigure}{\columnwidth}
%%    \includegraphics[width=\columnwidth]{%colPredShortTS}
%    \caption{\svdfive}
%    \label{fig:lmasvdfive}
%  \end{subfigure}%
%  \caption{{\color{red} [If Liz thinks we should %include these I actually need to generate this %figure]}An LMA-based forecast of
%       processor-efficiency performance trace from the%\col ,\gcc and \svdfive generating processes.  Red %circles and blue $\times$s are the true and
%       predicted values, respectively; vertical bars %show where these
%       values differ.}\label{fig:lmapredictions}
% \end{figure}
%
%\begin{figure}[htbp]
%  \centering
%   \includegraphics[width=\textwidth]{colPredShortTS}
%    \caption{A forecast of the last 4,000 points of the %signal in
%      Figure~\ref{fig:ipc} using an LMA-based strategy on %the
%      embedding in Figure~\ref{fig:embedding}.  Red circles %and blue
%      $\times$s are the true and predicted values, %respectively;
%      vertical bars show where these values differ.}
%\label{fig:lmacol}
%\end{figure}
%
%\begin{figure}[htbp]
%  \centering
%    \includegraphics[width=\textwidth]{colPredShortTS}
%     \caption{{\color{red} [actually need to generate this %figure]}An LMA-based forecast of the last 4,000 points of a
%       processor-load performance trace from the \gcc
%       benchmark.  Red circles and blue $\times$s are the %true and
%       predicted values, respectively; vertical bars show %where these
%       values differ.}
%\label{fig:lmagcc}
%\end{figure}
%
%On the other hand, if we use the same approach to %forecast the
%processor efficiency (IPC) of the \gcc time series, %the
%prediction is far less accurate; see Figure~%\ref{fig:lmagcc}.
%Figure~\ref{fig:lmasvdfive} is clearly a better %prediction than Figure~\ref{fig:gccLMA} but not nearly as good as the \col forecast.

### 2014-07-16 related.tex
second round of LizB edits

@@

>>>>

The notion of complexity used here differs from that considered by
e.g. \cite{Shalizi2008}, which would consider a time series without
any statistical regularities to be non-complex.

@@

As a final note: there has been prior work under a similar title to
ours~\cite{haven2005}, but there are only superficial similarities
between the two research projects.  Haven \emph{et al.} utilize the
relative entropy to quantify the difference in predictability between
two distributions: one evolved from a small ensemble of past states
using the known dynamical system, and the other the observed
distribution. Our work quantifies the predictability of a single
observed time series using weighted permutation entropy and makes no
assumptions about the generating process.  More closely related is the
work by Boffetta et al.~\cite{boffetta02}.  \alert{[[Sentences from
      Ryan explaining exactly what the difference is.]]]
